---
title: "damage_functions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
 Script to generate a damage function for the social cost of ocean acidification
 Corinne Hartin CCD/CSIB
 May 2022

 Inputs:
 
 timeseries of the following:

1. welfare estimates in USD for all scenarios
 
2. ocean pH global mean
 
3. global mean surface temperature

 Fit a function to both pH and GMT. Testing out Canada and USA for RCP2.6 and RCP8.5

```{r libraries, include = FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(equatiomatic)
library(plot3D)
```

```{r wlefare, echo = FALSE, include = FALSE}

# read in welfare estimates
welfare_us <- read_excel("data/US total annual economic impacts.xlsx") %>% 
  select(year, "RCP2.6 vs Present", "RCP8.5 vs Present") %>% 
  rename(rcp26 = "RCP2.6 vs Present") %>% 
  rename(rcp85 = "RCP8.5 vs Present") %>% 
  pivot_longer(cols = c("rcp26", "rcp85"),
               names_to = "scenario",
               values_to = "welfare"
                 ) %>% 
  mutate(welfare = welfare / 10^6) %>%  #convert to millions of dollars 
  mutate(region = "USA") %>% 
  as.data.frame()

welfare_can <- read_excel("data/Canada total annual economic impacts.xlsx") %>% 
  select(Year, "RCP2.6 vs Present", "RCP8.5 vs Present") %>% 
  rename(rcp26 = "RCP2.6 vs Present") %>% 
  rename(rcp85 = "RCP8.5 vs Present") %>%
  rename(year = "Year") %>% 
  pivot_longer(cols = c("rcp26", "rcp85"),
               names_to = "scenario",
               values_to = "welfare"
  ) %>% 
  mutate(welfare = welfare / 10^6) %>%  #convert to millions of dollars 
  mutate(region = "Canada") %>% 
  as.data.frame()

welfare <- full_join(welfare_us, welfare_can)
  

```
## Plot Welfare

```{r welfare, echo=FALSE}
ggplot(welfare, 
       aes(year, welfare, color = scenario)) +
  facet_grid(~ region, scales = "free") +
  geom_point()+
  theme_minimal()
  
```

```{r pH data, echo = FALSE}
# read in pH data
# Travis recommends taking the mean of the 3 models - pH values are both surface and bottom.  Do I just want surface? yes, b/c that's all the SCM can produce
# Or do i want to try with [CO2] - maybe try both
# global hydrogen ion - need to convert to pH

ph <- read.csv("data/GlobalYearlyAveragepH.csv") %>% 
  select(Year, GFDL.ESM2G.26hsurf, IPSL.CM5AMR.26hsurf, GFDL.ESM2G.85hsurf,
         IPSL.CM5AMR.85hsurf, MPI.ESMMR.26hsurf, MPI.ESMMR.85hsurf) %>% 
  pivot_longer(cols = c("GFDL.ESM2G.26hsurf", "IPSL.CM5AMR.26hsurf", "GFDL.ESM2G.85hsurf"
                        ,"IPSL.CM5AMR.85hsurf", "MPI.ESMMR.26hsurf", "MPI.ESMMR.85hsurf"),
               names_to = "model",
               values_to = "value") %>% 
  rename(year = Year) %>% 
  separate(model, c("model","B", "scenario") ) %>% 
  select(-B) %>% 
  pivot_wider(names_from = scenario, 
              values_from = value) %>% 
  group_by(year) %>% 
  summarise(rcp26 = mean(`26hsurf`), rcp85 = mean(`85hsurf`) ) %>% 
  mutate(rcp26 = -log10(rcp26*10^-6)) %>%
#    mutate(rcp26 = (rcp26*10^-6)) %>% 
  mutate(rcp85 = -log10(rcp85*10^-6)) %>% 
#  mutate(rcp85 = (rcp85*10^-6)) %>% 
  select(year, rcp26, rcp85) %>% 
  pivot_longer(cols = c("rcp26","rcp85"),
               names_to = "scenario",
               values_to = "pH")

avg_ph <- ph %>% 
  filter(year %in% c(2011:2020)) %>% 
  summarise(avg = mean(pH))

ph <- ph %>% 
  mutate("pH" = pH - avg_ph$avg) %>% 
  filter(year >2019) %>% 
  as.data.frame()
```
## Plot pH

```{r pH plot, echo = FALSE}
ggplot(ph, aes(year, pH, color = scenario))+
  geom_point()+
  theme_minimal()
```


```{r temperature data, echo = FALSE}
# temperature files from Hector comparison data

RCP26 <- read.csv("data/summary_rcp26.csv") %>% 
  filter(vtag == "tgav") %>% 
  select(year, meanvalue, scenario)

avg_26 <- RCP26 %>% 
  filter(year %in% c(2011:2020)) %>% 
  summarise(avg = mean(meanvalue))

RCP26 <- RCP26 %>% 
  mutate("temp" = meanvalue - avg_26$avg)

RCP85 <- read.csv("data/summary_rcp85.csv") %>% 
  filter(vtag == "tgav") %>% 
  select(year, meanvalue, scenario)

avg_85 <- RCP85 %>% 
  filter(year %in% c(2011:2020)) %>% 
  summarise(avg = mean(meanvalue))

RCP85 <- RCP85 %>% 
  mutate("temp" = meanvalue - avg_85$avg)

temperature <- rbind(RCP26, RCP85) %>% 
  filter(year %in% (2020:2100) ) %>% 
  select(-meanvalue)
```

## Plot Temperature

```{r plot temperature, echo = FALSE}
ggplot(temperature, aes(year, temp, color = scenario)) +
  geom_point() +
  theme_minimal()
```

```{r join data, echo = FALSE, include = FALSE}
canada <- left_join(welfare_can, temperature)
canada <- left_join(canada, ph)

usa <- left_join(welfare_us, temperature)
usa <- left_join(usa, ph)

total_welfare <- rbind(usa, canada)
```

## Plot Damages v Temp

```{r dam v temp, echo = FALSE}
ggplot(total_welfare, aes(temp, welfare, color = scenario))+
  geom_point() +
  facet_wrap(~region)+
  theme_minimal()
```

## Plot Damages v pH

```{r dam v pH, echo = FALSE}
ggplot(total_welfare, aes(pH, welfare, color = scenario))+
  geom_point() +
  facet_wrap(~region)+
  theme_minimal()
```

## test out different fits
### Canada
the small p-value below shows that the added complexity increases the fit, however, it looks like going to a polynomial fit does not increase the fit much more...

```{r anova can, echo = FALSE}
df <- total_welfare %>% 
  filter(region == "Canada")
  
model1 = lm(welfare~ temp, data=df )
model2 = lm(welfare~pH+temp, data = df)
model3 = lm(welfare~pH+poly(temp,2), data = df)

print(anova(model1, model2, model3))
#The most important value in the entire output is the p-value because this tells us whether there is a significant difference in the mean values between the three groups.
summary(model2)
extract_eq(model2, use_coefs = TRUE, fix_signs = FALSE)

summary(model3)
extract_eq(model3, use_coefs = TRUE, fix_signs = FALSE)
```

### USA
the small p-value below shows that the added complexity increases the fit, however, it looks like going to a polynomial fit does not increase the fit much more...

```{r anova usa, echo = FALSE}
df <- total_welfare %>% 
  filter(region == "USA")
  
model1 = lm(welfare~ temp, data=df )
model2 = lm(welfare~pH+temp, data = df)
model3 = lm(welfare~pH+poly(temp,2), data = df)

print(anova(model1, model2, model3))
#The most important value in the entire output is the p-value because this tells us whether there is a significant difference in the mean values between the three groups.
summary(model2)
extract_eq(model2, use_coefs = TRUE, fix_signs = FALSE)

summary(model3)
extract_eq(model3, use_coefs = TRUE, fix_signs = FALSE)
```
